---
title: "Luminiscience"
output: html_document
date: UPR Health Innovation Replicathon 2020
author: Corinne Croslyn, Omaris Vélez
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, message = FALSE)
```


## Reproducing the findings of the CCLE and GDSC Studies




### Exploratory analysis of pharmacogenomic data

```{r Loading Use Packages, include=FALSE}
library(ggplot2)
library(gapminder)
library(tidyverse)
theme_set(theme_bw())
library(energy)
```

```{r readingRawdata}
rawPharmacoData <- readRDS("~/OneDrive - University of Evansville/2019-2020/Summer 2020/IQ BIO/PR2020Replicathon/data/rawPharmacoData.rds")

modelSummarizedPharmacoData <- readRDS("~/OneDrive - University of Evansville/2019-2020/Summer 2020/IQ BIO/PR2020Replicathon/data/modelSummarizedPharmacoData.rds")
```




#### How many cell-lines are contained in the data?
```{r count-uniq, include=FALSE}
length(unique(rawPharmacoData$cellLine))
length(unique(rawPharmacoData$drug))
```


There are 288 cell lines and 15 drugs. 




#### How many drug concentrations were used in each study?
```{r count-conc, include=FALSE}
tapply(rawPharmacoData$concentration, rawPharmacoData$study,
       function(x) { length(unique(x)) })
```


There were 8 drug concentrations in study CCLE, and 32 drug concentrations in study GDSC.

```{r conc-hist, fig.width = 8, fig.height = 4, warning = FALSE}
rawPharmacoData %>%
    ggplot(aes(x = log2(concentration))) +
    geom_histogram(fill = "gray", color = "black") +
    facet_wrap(~ study) +
    ggtitle("Distributions of concentrations by study")
```




Based on the distributions of concentrations, the CCLE study seems much more consistent in their experimental protocol than the GDSC study when the drug concentrations are plotted. 




#### Histograms, apart from telling us how the data is distributed, can also make evident potential problems with the data. Plot a histogram of drug viabilities. Does it look as one would expect from the description of the data?
```{r log-density, fig.width = 8, fig.height = 4, warning = FALSE}
rawPharmacoData %>%
    ggplot(aes(x = viability, group = study, fill = study, color = study)) +
    geom_density(alpha = 1/4) +
    xlim(0,170) +
    ggtitle("Distributions of viability scores by study")
```



#### How many viability data points are within the expected range according to the definition of viability (e.g. above 0 and below 100)? Try to come up with explanations about the values that are out of range. Are these due to technical issues? Biology?
```{r viability-summaries, include=FALSE}
range(rawPharmacoData$viability)
too_small <- sum(rawPharmacoData$viability < 0)
too_small
too_big <- sum(rawPharmacoData$viability > 100)
too_big
not_viable <- too_big+too_small
not_viable
43427-not_viable
```



To be considered viable, cells have to have a viability of >0 but <100. In the raw pharmacodata set, viability ranges from -20 to 319.4919. There are a total of 15,801 data points that are out of the range of viability. Therefore, only 27,626 data points fall in the range of acceptable viability. Some cells might have less viability than others due natural mutations that could arise in the culture, "natural" sensitivity to the drug, or if the cell line wasn't kept in optimal conditions.


```{r viability-boxplots, fig.width = 8, fig.height = 4, warning = FALSE}
rawPharmacoData %>%
    ggplot(aes(y = viability, x = drug, fill = study)) +
    scale_x_discrete() + 
    annotate(geom = "rect", ymin = 0, ymax = 100, xmin = -Inf, xmax = Inf,
             fill = 'black', alpha = 1/6) +
    geom_boxplot(outlier.alpha = 1/5) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1/2)) +
    ggtitle("Distributions of viability scores by drug and study")
```




#### Read the csv file containing the summarized data. What kinds of variables are in the data? What does each column represent?
```{r readingSummarizedData}
summarizedPharmacoData <- readRDS("~/OneDrive - University of Evansville/2019-2020/Summer 2020/IQ BIO/PR2020Replicathon/data/summarizedPharmacoData.rds")
```

```{r exploringSummarized, include=FALSE}
head(summarizedPharmacoData)
str(summarizedPharmacoData)
```


There are 6 varibles in the summarized data set: cellLine, drug, ic50_CCLE, auc_CCLE, ic50_GDSC, and auc_GDSC. Each column in the dataset is a different variable. 




#### Plot histograms of the viability scores at different levels of drug doses. Are stronger drug concentrations consistent with lower viability scores?
```{r viabilityRawData, fig.width=4, fig.height=8}
rawPharmacoData %>%
    ggplot(aes(x =viability)) +
    geom_histogram(fill = "gray", color = "black") +
    facet_wrap(~ cut_number(concentration, n=5), ncol = 1) +
    ggtitle("Distributions of viability by concentration")
```

```{r moreExploration, fig.width=4, fig.height=8}
rawPharmacoData %>%
    ggplot(aes(x =viability)) +
    geom_histogram(fill = "gray", color = "black") +
    facet_grid(cut_number(concentration, n=5)~study) +
    ggtitle("Distributions of viability by concentration")
```




### Using Correlation Measures to Assess Replicability of Drug Response Studies




#### Create a scatterplot for each drug comparing the AUC in GDSC and CCLE for all cell lines.
```{r auc-all-drug-summarized, fig.width = 10, fig.height = 9}
summarizedPharmacoData %>%
    ggplot(aes(x = auc_GDSC, y = auc_CCLE)) +
    geom_point(alpha = 1/2, cex = 1/2) + 
    facet_wrap(~ drug) +
    xlab("GDSC AUC") +
    ylab("CCLE AUC") +
    ggtitle("AUC summaries of cell line response across studies (Summerized Data)")
```

```{r ic50-all-drug-scaled-summarized, fig.width = 10, fig.height = 9}
summarizedPharmacoData %>%
    ggplot(aes(x = -log10(ic50_GDSC / 10^6),
               y = -log10(ic50_CCLE / 10^6))) +
    geom_point(cex = 1/2) + 
    facet_wrap(~ drug) +
    xlab("-log10(GDSC IC50/10^6)") +
    ylab("-log10(CCLE IC50/10^6)") +
    ggtitle("IC50 summaries of cell line response across studies (Summarized Data)")
```

```{r auc-all-drug-model, fig.width = 10, fig.height = 9, warning=FALSE}
modelSummarizedPharmacoData %>%
    ggplot(aes(x = auc_GDSC, y = auc_CCLE)) +
    geom_point(alpha = 1/2, cex = 1/2) + 
    facet_wrap(~ drug) +
    xlab("GDSC AUC") +
    ylab("CCLE AUC") +
    ggtitle("AUC summaries of cell line response across studies (Model Data)")
```


```{r ic50-all-drug-scaled-model, fig.width = 10, fig.height = 9, warning=FALSE}
modelSummarizedPharmacoData %>%
   ggplot(aes(x = -log10(ic50_GDSC / 10^6),
              y = -log10(ic50_CCLE / 10^6))) +
    geom_point(alpha= 1/2, cex = 1/2) + 
    facet_wrap(~ drug) +
    xlab("GDSC IC50") +
    ylab("CCLE IC50") +
    ggtitle("IC50 summaries of cell line response across studies (Model Data)")
```


#### Calculate correlation coefficients of the AUC in GDSC and CCLE for each drug  

```{r auc-correlation-summarized, fig.width=8, fig.height=4}
AUCdrugCorrs <- summarizedPharmacoData %>% 
    filter(!is.na(auc_GDSC), !is.na(auc_CCLE)) %>% 
    group_by(drug) %>%
    summarize(Pearson_auc  = cor(auc_GDSC, auc_CCLE, method = "pearson"),
              Spearman_auc = cor(auc_GDSC, auc_CCLE, method = "spearman"),
               dcor_auc = dcor(auc_GDSC, auc_CCLE,index = 1))

AUCdrugCorrs <- gather(AUCdrugCorrs, measure, correlation, -drug)

AUCdrugCorrs %>%
    ggplot(aes(x = drug, y = correlation, fill = measure, group = measure)) +
    geom_bar(stat = "identity", position = position_dodge(), colour = "black") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_grey() +
    ylim(-.1, 1) + 
    ggtitle("Correlation of cell line AUC summaries between studies for each drug (Summarized Data)")
```

```{r ic50-correlation-summarized, fig.width=8, fig.height=4}
IC50drugCorrs <- summarizedPharmacoData %>% 
    group_by(drug) %>%
    filter(!is.na(ic50_GDSC), !is.na(ic50_CCLE)) %>% 
    summarize(Pearson_ic50  = cor(-log10(ic50_GDSC / 10^6), -log10(ic50_CCLE / 10^6), method = "pearson"),
              Spearman_ic50 = cor(-log10(ic50_GDSC / 10^6), -log10(ic50_CCLE / 10^6), method = "spearman"),
              dcor_ic50 = dcor(-log10(ic50_GDSC / 10^6),
                                     -log10(ic50_CCLE / 10^6),
                                     index = 1))

IC50drugCorrs <- gather(IC50drugCorrs, measure, correlation, -drug)

IC50drugCorrs %>%
    ggplot(aes(x = drug, y = correlation, fill = measure, group = measure)) +
    geom_bar(stat = "identity", position = position_dodge(), colour = "black") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_grey() +
    ylim(0, 1) + 
    ggtitle("Correlation of cell line IC50 summaries between studies for each drug (Summarized Data)")
```

```{r auc-correlation-model, fig.width=8, fig.height=4}
AUCdrugCorrsmodel <- modelSummarizedPharmacoData %>% 
     filter(!is.na(auc_GDSC), !is.na(auc_CCLE)) %>%
    group_by(drug) %>%
    summarize(Pearson_auc_model  = cor(auc_GDSC, auc_CCLE, method = "pearson", use="na.or.complete"),
              Spearman_auc_model = cor(auc_GDSC, auc_CCLE, method = "spearman", use="na.or.complete"),
              dcor_auc_model = dcor(auc_GDSC,auc_CCLE,index = 1))

AUCdrugCorrsmodel <- gather(AUCdrugCorrsmodel, measure, correlation, -drug)

AUCdrugCorrsmodel %>%
    ggplot(aes(x = drug, y = correlation, fill = measure, group = measure)) +
    geom_bar(stat = "identity", position = position_dodge(), colour = "black") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_grey() +
    ylim(-.15, 1) + 
    ggtitle("Correlation of cell line AUC summaries between studies for each drug (Model Data)")
```

```{r ic50-correlation-model, fig.width=8, fig.height=4}
IC50drugCorrsmodel <- modelSummarizedPharmacoData %>% 
    filter(!is.na(ic50_GDSC), !is.na(ic50_CCLE)) %>% 
    group_by(drug) %>%
    summarize(Pearson_ic50_model  = cor(-log10(ic50_GDSC / 10^6), -log10(ic50_CCLE / 10^6), method = "pearson",use="na.or.complete"),
              Spearman_ic50_model = cor(-log10(ic50_GDSC / 10^6), -log10(ic50_CCLE / 10^6), method = "spearman",use="na.or.complete"),
              dcor_ic50_model = dcor(-log10(ic50_GDSC / 10^6),
                                     -log10(ic50_CCLE / 10^6),
                                     index = 1))

IC50drugCorrsmodel <- gather(IC50drugCorrsmodel, measure, correlation, -drug)


IC50drugCorrsmodel %>%
    ggplot(aes(x = drug, y = correlation, fill = measure, group = measure)) +
    geom_bar(stat = "identity", position = position_dodge(), colour = "black") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_grey() +
    ylim(-.10, 1) + 
    ggtitle("Correlation of cell line IC50 summaries between studies for each drug (Model Data)")
```



#### Based on the scatterplot and correlation values, would you say that they tend to agree? Why or why not?


They don't tend to agree because the scatterplot and correlation values aren't consistent, especially when you compare the Pearson and the Spearman. The correlations also are not very strong, with few being above .70. Many of the correlations show moderate correlation, falling between .20-.70. 




#### Does the AUC or IC50 suggest more agreement between the studies?


In the summarized data, the AUC seems to suggest more agreement than the IC50. Overall, the correlation of AUC seems to be stronger then the IC50, with IC50 having a greater number of weaker correlations (correlations <.25).This could be because the model that IC50 used to predict when cells would reach 50% viability is a bad predictor. 




#### Which drug do you think shows the most consistency between the studies? How about the least? 


17-AAG seems to be the most consistent across different correlation measures and the scatterplots. Nutilnib seems to be the least consistent across different correlation measures and scatterplots.  




#### If you calculated more than one type of correlation coefficient (for example Pearson and Spearman), how do they differ? Which do you think is a better summary of the consistency between the two studies?


The Pearson correlation coefficient measures the degree of linear between variables, while Spearman's correlation coefficient measures the agreement of the rankings between variables. The distance correlation coefficient measures both linear and nonlinear association between two variables. For interval scale, the Pearson correlation coefficient is typically more appropriate. Therefore, we believe that the Pearson coefficient is more appropriate measure of correlation for this data. Additionally, the Pearson and the distance correlation coefficients tend to agree with one another even when the Spearman's is vastly different. This supports the use of Pearson's correlation coefficient as a better measure of correlation for this data.  


